---
title: "Problem Set 1"
author: "Aaron Graybill"
output: pdf_document
header-includes:
  - \usepackage{amsmath}

---

I did not collaborate with anyone on any of these questions. Also the spacing is
very weird because of Rmarkdown, sorry about that!

# Problem 1

## 1.a 

Intuitively, to have a unique solution, we must have that both $b_0$ and $b_1$
actually change the objective function. For example, notice that if $x_i=0$ for
all $i$, then the choice of $b_1$ is completely immaterial. Similarly if $x_i=c$
for a constant $c$ for all $i$, changing $b_1$ has the same effect to changing $b_0$
so there would be a multiplicity of solutions. Therefore, we get a unqiue solution
when there is variance in the $x_i$s. If there were more variables, this equivalent
to requiring the design matrix to be full rank.

## 1.b

A nice way to derive the desired condition is to first define $b=\begin{pmatrix} b_0 \\ b_1 \end{pmatrix}$. Then we express the optimization problem as:

$$
\begin{aligned}
\arg\min_b \sum_i\left(y_i-b'z_i\right)^2
\end{aligned}
$$

Which has the (vector) first order condition:

$$
\begin{aligned}
\sum_i 2(y_i-b'z_i)(-z_i')&=0\\
2\sum_i (y_i-b'z_i)(-z_i')&=0\\
\sum_i (y_i-b'z_i)(-z_i')&=0\\
\sum_i -y_iz_i'+b'z_iz_i'&=0\\
\sum_i b'z_iz_i'&=\sum_iy_iz_i'\\
b'\sum_i z_iz_i'&=\sum_iy_iz_i'\\
b'&=\left(\sum_iy_iz_i'\right)\left(\sum_i z_iz_i'\right)^{-1}\\
b&=\left(\left(\sum_iy_iz_i'\right)\left(\sum_i z_iz_i'\right)^{-1}\right)'\\
b&=\left(\left(\sum_i z_iz_i'\right)^{-1}\right)'\left(\sum_iy_iz_i'\right)'\\
b&=\left(\left(\sum_i z_iz_i'\right)^{-1}\right)'\left(\sum_iz_iy_i\right)\\
b&=\left(\sum_i z_iz_i'\right)^{-1}\left(\sum_iz_iy_i\right)\\
\end{aligned}
$$


That last line uses the fact that the inverse of that sum-of-outer-product matrices
must be symmetric since it is the sum of symmetric matrices, so its inverse
must be symmetric.

## 1.c

One way to approach this question is to rederive $\hat\beta$ as though they were
scalars. Taking the derivative wrt $b_0$, the FOC would be:

$$
\begin{aligned}
\sum_i2(y_i-b_0-b_1x_i)=0\\
\sum_i(y_i-b_0-b_1x_i)=0\\
\sum_i(y_i-b_0-b_1x_i)=0\\
-nb_0+\sum_iy_i-b_1x_i&=0\\
\sum_iy_i-b_1x_i&=nb_0\\
\frac{1}{n}\sum_iy_i-b_1x_i&=b_0\\
\end{aligned}
$$

Then the FOC for $b_1$ is:

$$
\begin{aligned}
\sum_i2(y_i-b_0-b_1x_i)(-x_i)&=0\\
\sum_i(y_i-b_0-b_1x_i)(-x_i)&=0\\
\sum_i(-y_ix_i+b_0x_i+b_1x_i^2)&=0\\
\sum_ib_0x_i&=\sum_i y_ix_i-b_1x_i^2\\
b_0\sum_ix_i&=\sum_i y_ix_i-b_1x_i^2\\
\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\sum_ix_i&=\sum_i y_ix_i-b_1x_i^2\\
\frac{1}{n}\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\sum_ix_i&=\frac{1}{n}\sum_i y_ix_i-b_1x_i^2\\
\frac{1}{n}\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\sum_ix_i&=\frac{1}{n}\sum_i y_ix_i-\frac{1}{n}\sum_ib_1x_i^2\\
\frac{1}{n}\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\sum_ix_i&=\frac{1}{n}\sum_i y_ix_i-b_1\frac{1}{n}\sum_ix_i^2\\
\frac{1}{n}\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\sum_ix_i&=\frac{1}{n}\sum_i y_ix_i-b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)*\\
\frac{1}{n}\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\sum_ix_i&=(\mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y])-b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)*\\
\left(\frac{1}{n}\sum_iy_i-b_1x_i\right)\mathbb{E}_n[x]&=(\mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y])-b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)\\
\left(\mathbb{E}_n[y]-b_1\mathbb{E}_n[x]\right)\mathbb{E}_n[x]&=(\mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y])-b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)\\
\left(\mathbb{E}_n[y]-b_1\mathbb{E}_n[x]\right)\mathbb{E}_n[x]+b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)&=(\mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y])\\
\mathbb{E}_n[y]\mathbb{E}_n[x]-b_1\mathbb{E}_n[x]^2+b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)&=(\mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y])\\
-b_1\mathbb{E}_n[x]^2+b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)&=(\mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y])-\mathbb{E}_n[y]\mathbb{E}_n[x]\\
-b_1\mathbb{E}_n[x]^2+b_1\left(\mathbb{V}_n[x]+\mathbb{E}_n[x]^2\right)&=\mathbb{C}_n[x,y]\\
b_1\mathbb{V}_n[x]&=\mathbb{C}_n[x,y]\\
b_1&=\mathbb{C}_n[x,y]/\mathbb{V}_n[x]\\
\end{aligned}
$$

Throughout that proof, I make extensive use of the definitions of $\mathbb{E}$. In
the lines with $*$, I make use of a useful decomposition of the covariance (and by proxy the variance):

$$
\begin{aligned}
\mathbb{C}\text{ov}_n(x,y)&\equiv\frac{1}{n}\sum_i(x_i-\mathbb{E}_n[x])(y_i-\mathbb{E}_n[y])\\
&=\frac{1}{n}\sum_i x_iy_i-\frac{1}{n}\mathbb{E}_n[x]\sum_iy_i-\frac{1}{n}\mathbb{E}_n[x]\sum_i y_i+\sum_i\frac{1}{n}\mathbb{E}_n[x]\mathbb{E}_n[y]\\
&=\frac{1}{n}\sum_i x_iy_i-\mathbb{E}_n[x]\mathbb{E}_n[y]-\mathbb{E}_n[x]\mathbb{E}_n[x]+\mathbb{E}_n[x]\mathbb{E}_n[y]\\
&=\frac{1}{n}\sum_i x_iy_i-\mathbb{E}_n[x]\mathbb{E}_n[y]
\end{aligned}
$$

In particular, I use, $\frac{1}{n}\sum xy = \mathbb{C}_n[x,y]+\mathbb{E}_n[x]\mathbb{E}_n[y]$
for both the variance and the covariance.

Now plugging into the FOC for $b_0$ we can recover $\hat\beta_0$:

$$
\begin{aligned}
b_0&=\frac{1}{n}\sum_iy_i-b_1x_i\\
&= \mathbb{E}_n[y]-\frac{\mathbb{C}_n[x,y]}{\mathbb{V}_n[x]} \mathbb{E}_n[x]
\end{aligned}
$$




## 1.d

I apologize for how meandering this proof is.

Note first that when $x$ is binary, we must have:

$$
\mathbb{E}_n[x]\equiv\frac{1}{n}\sum_{i=1}x_i=\frac{1}{n}\sum_{i:x_i=1} 1+\frac{1}{n}\sum_{i:x_i=0} 0=\frac{1}{n}\sum_{i:x_i=1} 1
$$

Recall the covariance decomposition: $\frac{1}{n}\sum_i x_iy_i-\mathbb{E}_n[x]\mathbb{E}_n[y]$

When $x_i$ is binary, then the first part of the sum there only includes the $y_i$
where $x_i=1$. Which is to say that the covariance is simply:

$$
\frac{1}{n}\sum_{x_i=1} y_i-\mathbb{E}_n[x]\mathbb{E}_n[y]
$$

And notice that $\mathbb{E}_n[y]$ could alternately be written as:

$$
\begin{aligned}
\mathbb{E}_n[y]&\equiv\frac{1}{n}\sum_{i} y_i
\\&=\frac{1}{n}\left(\sum_{x_i=0} y_i + \sum_{x_i=1}y_i\right)
\end{aligned}
$$

Which means the covariance is:

$$
\begin{aligned}
\mathbb{C}\text{ov}_n(x,y)&=\frac{1}{n}\sum_{x_i=1} y_i-\mathbb{E}_n[x]\frac{1}{n}\left(\sum_{x_i=0} y_i + \sum_{x_i=1}y_i\right)\\
&=\frac{1}{n}\left(\sum_{x_i=1} y_i-\mathbb{E}_n[x]\left(\sum_{x_i=0} y_i + \sum_{x_i=1}y_i\right)\right)\\
&=\frac{1}{n}\left((1-\mathbb{E}_n[x])\sum_{x_i=1} y_i-\mathbb{E}_n[x]\sum_{x_i=0} y_i \right)\\
&=(1-\mathbb{E}_n[x])\frac{1}{n}\sum_{x_i=1} y_i-\mathbb{E}_n[x]\frac{1}{n}\sum_{x_i=0} y_i
\end{aligned}
$$

Using the simplification of the covariance, $\frac{1}{n}\sum_i x_iy_i-\mathbb{E}_n[x]\mathbb{E}_n[y]$ to compute the variance, we get:

$$
\begin{aligned}
\frac{1}{n}\sum_{x_i=1}1 -\frac{1}{n}\sum_{x_i=1}1\frac{1}{n}\sum_{x_i=1}1\\
\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(1-\frac{1}{n}\sum_{x_i=1}1\right)\\
\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(\frac{1}{n}\sum_{x_i=0}1\right)
\end{aligned}
$$

Where the last line uses the property that $\sum_{x_i=0}1 + \sum_{x_i=1}1=n$.

Dividing variance and covariance, we have:

$$
\begin{aligned}
\hat\beta_1&=\frac{(1-\mathbb{E}_n[x])\frac{1}{n}\sum_{x_i=1} y_i-\mathbb{E}_n[x]\frac{1}{n}\sum_{x_i=0} y_i}{\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(\frac{1}{n}\sum_{x_i=0}1\right)}\\
&=\frac{(1-\mathbb{E}_n[x])\frac{1}{n}\sum_{x_i=1} y_i}{\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(\frac{1}{n}\sum_{x_i=0}1\right)}-\frac{\mathbb{E}_n[x]\frac{1}{n}\sum_{x_i=0} y_i}{{\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(\frac{1}{n}\sum_{x_i=0}1\right)}}\\
&=\frac{\left(\frac{1}{n}\sum_{x_i=0}1\right)\frac{1}{n}\sum_{x_i=1} y_i}{\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(\frac{1}{n}\sum_{x_i=0}1\right)}-\frac{\left(\frac{1}{n}\sum_{x_i=1}1\right)\frac{1}{n}\sum_{x_i=0} y_i}{{\left(\frac{1}{n}\sum_{x_i=1}1\right)\left(\frac{1}{n}\sum_{x_i=0}1\right)}}\\
&=\frac{\frac{1}{n}\sum_{x_i=1} y_i}{\frac{1}{n}\sum_{x_i=1}1}-\frac{\frac{1}{n}\sum_{x_i=0} y_i}{{\frac{1}{n}\sum_{x_i=0}1}}\\
&=\frac{\sum_{x_i=1} y_i}{\sum_{x_i=1}1}-\frac{\sum_{x_i=0} y_i}{{\sum_{x_i=0}1}} \quad \square\\
\end{aligned}
$$

\newpage

# Problem 2

```{r}
data = read.csv('nls.csv')
y <- data$luwe
x <- data$educ
ones <- rep(1, length(x))

cov <- function(y,x){
  # implicitly this takes care of 1/n
  mean(x*y)-(mean(x)*mean(y))
}

beta_1 <- cov(y,x)/cov(x,x)
beta_0 <- mean(y)-(beta_1*mean(x))
```

```{r, echo = F}
print(paste0("My Estimated slope is: ",beta_1))
print(paste0("My Estimated intercept is: ",beta_0))
```

Here I compare my calculations to the in-built function:
```{r}
benchmark_coefs <- lm(y ~ x)$coefficients |> unname()

all.equal(benchmark_coefs,c(beta_0,beta_1))
```
