---
title: "Problem Set 5"
author: 
  - John Cho^[Authors listed alphabetically by and contributed equally to every question.]
  - Aaron Graybill
  - Thomas Li
  - Blair Moore
output: pdf_document
header-includes:
  - \usepackage{amsmath}
---

```{r}
seed_string <- "Blair, Thomas, Aaron, John"
seed_number <- 
  seed_string |>  
  rlang::hash() |> # hash the string to a hex number
  substr(1, 8) |> # make the number smaller 
  strtoi(base = 16) # convert to decimal

set.seed(seed_number)
```

# Hamilton's Problems

## Exercise 3.1

## Exercise 3.2

We can rewrite $Y_t$ in the form 
$$\xi_t = \mathbf{F} \xi_{t - 1} + \eta_t.$$
Then,
$$\begin{aligned}
\begin{bmatrix}
y_t \\
y_{t-1}
\end{bmatrix} &= \begin{bmatrix}
1.1 & -0.18 \\
1 & 0
\end{bmatrix} \begin{bmatrix}
y_{t-1} \\
y_{t-2}
\end{bmatrix} + \begin{bmatrix}
\varepsilon_t \\
0
\end{bmatrix}
\end{aligned}$$

We find the eigenvalues of $\mathbf{F}$ by solving
$$\begin{aligned}
\det \begin{pmatrix} 1.1 - \lambda & -0.18 \\ 1 &-\lambda\end{pmatrix} &=  0 \\
-1.1 \lambda + \lambda^2 + 0.18 &= 0\\ 
\lambda^2 - 1.1 \lambda + 0.18 &=  0 \\ 
(\lambda - 0.9)(\lambda - 0.2) &=  0.
\end{aligned}$$

Therefore, the two eigenvalues are $\lambda_1 = 0.2$ and $\lambda_2 = 0.9$. Because 
$$max_i |\lambda_i| < 1$$
we know that this AR(2) process admits a covariance-stationary MA($\infty$) process.


To calculate the mean and autocovariance, we proceed with the eigendecomposition of
$\mathbf{F}$, which is:$$\begin{aligned}
\mathbf{T}\mathbf{\Lambda}\mathbf{T}^{-1} &= \begin{pmatrix} 0.2 & 0.9 \\ 1 & 1 \end{pmatrix} \begin{pmatrix} 0.2 & 0 \\ 0 & 0.9 \end{pmatrix} \begin{pmatrix} -\frac{10}{7} & \frac{9}{7} \\ \frac{10}{7} & -\frac{2}{7} \end{pmatrix}
\end{aligned}$$

Then,
$$\begin{aligned}
F_{1,1}^j &= 0.2^2\left(- \frac{10}{7}\right) + 0.9^2 \left( \frac{10}{7}\right)\\
&= \left(\frac{-2}{7}\right)0.2^j + \left(\frac{9}{7}\right)0.9^j.
\end{aligned}$$

If $\psi_j =  \left(-\frac{2}{7}\right)0.2^j + \left(\frac{9}{7}\right)0.9^j$, then
$$ Y_t = \sum_{j = 0}^{\infty} \psi_j \epsilon_{t-j}.$$

We can see that
$$\mathbb{E}[Y_t] = 0.$$

For the autocovariance, 
$$\begin{aligned}
\gamma_{j} &= \mathbb{E}[(Y_t - \mu)(Y_{t-k} - \mu)\\
&= \mathbb{E}(Y_{t} Y_{t-k}) = \mathbb{E} \left[ \left( \sum_{j=0}^{\infty} \psi_{j} \varepsilon_{t-j} \right) \left( \sum_{l=0}^{\infty} \psi_{l} \varepsilon_{t-l} \right) \right] \\
&= \mathbb{E} \left[ \sum_{n=0}^{\infty} \sum_{j=0}^{n} \left(\psi_{j}\varepsilon_{t-j}\right) \left(\psi_{n-j} \varepsilon_{t-(n-j)}\right) \right] \text{(by Cauchy product)}\\
&= \sum_{n=0}^{\infty} \sum_{j=0}^{n} \psi_{j} \psi_{n-j} \mathbb{E}[\varepsilon_{t-j} \varepsilon_{t-n+j}] \\
\end{aligned}$$

We know that $\mathbb{E}[\varepsilon_{t-j} \varepsilon_{t - n + j}] = 1$ when the indexes are equal to each other $t- j = t - n + j$, which occurs when $n = 2j$. Suppose $n = 0$, then $j = 0$ and we have $\psi_0 \psi_0$. Suppose $n = 2$, then $j = 1$ and we have $\psi_1 \psi_1$, etc. 

Thus, we can observe that the above expression simplifies to 
$$\begin{aligned}
\gamma_j &= \sum_{j=0}^{\infty} \psi_{j}^{2} \mathbf{1} \{j \text{ is even}\} \\
&= \sum_{j=0}^{\infty} \left[ \left( -\frac{2}{7} \right) 0.2^{j} + \left( \frac{9}{7} \right) 0.9^{j} \right]^{2} \mathbf{1} \{j \text{ is even}\} \\
\end{aligned}$$




## Excercise 3.3

## Excercise 11.5

# The Permanent Income Model

## 1. Finding the Autocovariances

## 2. The Vector Process

## 3. Simulation Exercise

# The Kalman Filter

## 1. Analytical Moments

We are given the ARMA(1,1) process:

$$
y_t = \phi y_{t-1} + \varepsilon_t+\gamma \varepsilon_{t-1}
$$

In lag operator notation, we have:

$$
\begin{aligned}
y_t &= \phi L y_{t} + \varepsilon_t+\gamma L \varepsilon_{t}\\
(1-\phi L)y_t &= (1+\gamma L) \varepsilon_{t}
\end{aligned}
$$

The multiplicative inverse of $(1-\phi L)$ is $(1+\phi L+\phi^2 L^2-\ldots)$, and 
so we can alternately express the ARMA process as an MA($\infty$):

$$
\begin{aligned}
(1+\phi L+\phi^2 L^2-\ldots)(1-\phi L)y_t &= (1+\phi L+\phi^2 L^2-\ldots)(1+\gamma L) \varepsilon_{t}\\
y_t&=(1+\phi L+\phi^2 L^2-\ldots)(1+\gamma L) \varepsilon_{t}\\
&=[(1+\phi L+\phi^2 L^2-\ldots) + (\gamma L+\gamma\phi L+\gamma\phi^2 L^2+\ldots)]\varepsilon_{t} \\
&=[\underbrace{1}_{\psi_0}+\underbrace{\phi^0(\phi+\gamma)}_{\psi_1}L-\underbrace{\phi^1(\phi+\gamma)}_{\psi_2}L^2-\underbrace{\phi^2(\phi+\gamma)}_{\psi_3}L^3+\ldots]\varepsilon_{t}\\
&=\psi(L)\varepsilon_t
\end{aligned}
$$

Assuming at every point that each step is well defined.

In this form, it is much easier to compute the various moments of the distribution.

Namely we have:

$$
\begin{aligned}
E[y_t]&=E\left[[1+\phi^0(\phi+\gamma)L+\phi^1(\phi+\gamma)L^2+\phi^2(\phi+\gamma)L^3-\ldots]\varepsilon_{t}\right]\\
&=[1+\phi^0(\phi+\gamma)L+\phi^1(\phi+\gamma)L^2+\phi^2(\phi+\gamma)L^3-\ldots]E[\varepsilon_{t}]\\
&=[1+\phi^0(\phi+\gamma)L+\phi^1(\phi+\gamma)L^2+\phi^2(\phi+\gamma)L^3-\ldots]0\\
&= 0
\end{aligned}
$$

The variance of any $MA(\infty)$ process is given by:

$$
\begin{aligned}
\gamma_0&=\sigma^2\sum_{j=0}^\infty \psi_j^2 \\
&=\sigma^2\left(\psi_0^2+\sum_{j=1}^\infty \psi_j^2\right) \\
&=\sigma^2\left(1^2+\sum_{j=1}^\infty (\phi^{j-1}(\phi+\gamma))^2\right)\\
&=\sigma^2\left(1+(\phi+\gamma))^2\sum_{j=1}^\infty {\phi^{j-1}}^2\right)\\
&=\sigma^2\left(1+(\phi+\gamma))^2\sum_{j=0}^\infty {\phi^j}^2\right)\\
&=\sigma^2\left(1+(\phi+\gamma))^2\sum_{j=0}^\infty {\phi^2}^j\right)\\
&=\boxed{\sigma^2\left(1+(\phi+\gamma))^2\frac{1}{1-\phi^2}\right)}\\
\end{aligned}
$$

requiring that $\phi^2<1\implies |\phi|<1$

The $j$th autocovariance of the MA($\infty$) is given by:

$$
\begin{aligned}
\gamma_j &= \sigma^2\sum_{k=0}^\infty\psi_k\psi_{j+k} \\
&=\sigma^2\left(\psi_0\psi_j + \sum_{k=1}^\infty \psi_k\psi_{j+k} \right) \\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +\sum_{k=1}^\infty \phi^{k-1}(\phi+\gamma)\phi^{j+k-1}(\phi+\gamma)\right) \\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +\sum_{\kappa=0}^\infty \phi^{\kappa}(\phi+\gamma)\phi^{j+\kappa}(\phi+\gamma)\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\sum_{\kappa=0}^\infty \phi^{\kappa}\phi^{j+\kappa}\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\sum_{\kappa=0}^\infty \phi^{2\kappa}\phi^{j}\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\phi^{j}\sum_{\kappa=0}^\infty \phi^{2\kappa}\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\phi^{j}\frac{1}{1-\phi^2}\right)\\
&=\sigma^2\phi^{j-1}(\phi+\gamma) \left(1 +(\phi+\gamma)\phi\frac{1}{1-\phi^2}\right)\\
&=\sigma^2\phi^{j-1}(\phi+\gamma) \left(1 +(\phi+\gamma)\frac{\phi}{1-\phi^2}\right)\\
&=\sigma^2\phi^{j-1}(\phi+\gamma) \left(\frac{{1-\phi^2}}{{1-\phi^2}} +\frac{\phi^2+\phi\gamma}{1-\phi^2}\right)\\
&=\boxed{\sigma^2\phi^{j-1}(\phi+\gamma) \left(\frac{{1+\phi\gamma}}{{1-\phi^2}}\right)}
\end{aligned}
$$

Truly a bizarre expression that leaves little room for intuition. However, the series
convergence there also only required that $\phi^2<1$, so $|\phi|<1$ is still sufficient to
ensure covariance stationarity.

## 2. Conditional Expectations

## 3. Conditional Likelihood

## 4. State Space Representation

A state space representation is a system of vector equations of the form:

$$
\begin{aligned}
y_t&=\mu+Az_t+H\eta_t\\
z_t&=\Phi z_{t-1}+R\xi_t
\end{aligned}
$$

Where $\eta_t\sim \mathcal{N}(0,I),\xi_t\sim\mathcal{N}(0,\Sigma)$. Letting $z_{t}=\begin{bmatrix} y_t & \varepsilon_t \end{bmatrix}'$ we can represent the observation equation as:

$$
y_t = \begin{bmatrix} 1 & 0 \end{bmatrix}\begin{bmatrix} y_t \\ \varepsilon_t\end{bmatrix}+0\eta_t
$$

The $\eta$ component will not be needed, so we can set $H=0$ . Here's a canonical
way to represent the state equation that ensures dimensional conformability at every step.

$$
\begin{bmatrix} y_t \\ \varepsilon_t \end{bmatrix} =
\begin{bmatrix} \phi&\gamma \\ 0&0
\end{bmatrix}
\begin{bmatrix} y_{t-1} \\ \varepsilon_{t-1} \\\end{bmatrix} +
\begin{bmatrix}1&0 \\ 1&0\end{bmatrix} \xi_t
$$

Where $\xi_t$ is a 2-long vector distributed $\xi_t\sim\mathcal{N}(0,\sigma^2 I)$.
You could also use a one dimensional $\xi_t$ and then just the first column of the
matrix before $\xi_t$.

## 5. Simulation

```{r}
capT = 150
# structural parameters
phi = .9
gamma = .1
sigma_squared = 1

# state space parameters
dim_state <- 2
Sigma = sigma_squared * diag(nrow = dim_state)
R = matrix(c(1, 0, 
             1, 0),
           byrow = T, nrow = dim_state)
Phi = matrix(c(gamma, phi,
               0, 0),
             byrow = T, nrow = dim_state)

# Observation parameters
dim_obs <- 1 #
mu <- c(0)
A <- matrix(c(1, 0), byrow = T, nrow = dim_obs, ncol = dim_state)
H <- matrix(c(0), byrow = T, nrow = dim_obs)

evolve_state <- function(z, Phi, R, Sigma){
  # compute multivariate normal using linearity
  xi = Sigma%*%rnorm(nrow(Sigma))
  
  (Phi %*% z) + R%*%xi
}

evolve_observations <- function(z, mu, A, H){
  eta <- rnorm(nrow(H))
  
  mu + (A %*% z) + H%*%eta
}

z0 = c(0,0)
simulate <- function(capT, z0, Phi, R, Sigma, mu, A, H){
  # loop capT times, feeding the result of the last iteration into
  # this this iteration
  states <- purrr::accumulate(
  1:capT,
  ~ evolve_state(.x, Phi, R, Sigma),
  .init = z0
)
  
  # compute observation from latent state
  observations <- lapply(states, function(z){
    evolve_observations(z, mu, A, H)
  })
  
  # return long matrices with one row per time step
  states <- do.call(cbind, states) |> t()
  observations = do.call(cbind, observations) |> t()
  
  
  return(list(states = states, observations = observations))
}

sim_results <- simulate(capT, c(0,0), Phi, R, Sigma, mu, A, H)

library(ggplot2)
ggplot()+
  geom_line(aes(x=0:(nrow(sim_results$observations)-1), 
                y=sim_results$observations[,1]))+
  labs(y = expression("y"[t]*""),
       x = "t")
```

## 6. Implementing the Kalman Filter

```{r}
predict_y <- function(z, mu, A){
  mu + (A %*% z)
}

compute_v <- function(y, z, mu, A){
 y - predict_y(z, mu, A)
}

v_posterior_variance <- function(P, A, H){
  (A %*% P %*% t(A)) + H%*%t(H)
}

prediction_error_variance <- function(z, P, A){
  E
}

```

## 7. Applying the Kalman Filter

# Metropolis Hastings

## 1. Numerical Maximization

## 2. Generating M-H Draws

## 3. Lots of Simulations