---
title: "Problem Set 5"
author: 
  - John Cho^[Authors listed alphabetically by and contributed equally to every question.]
  - Aaron Graybill
  - Thomas Li
  - Blair Moore
  - Angie Wang
  - Roshie Xing
output: pdf_document
header-includes:
  - \usepackage{amsmath}
---

### REMEMBER TO GO TO CHUNK "lots-of-simulation-be-patient" AND CHANGE N TO 500000

```{r setup, include=FALSE}
rm(list=ls())  # clear environment
library(mvtnorm)
library(dplyr)
library(tidyr)
library(ggplot2)
```

```{r}
seed_string <- "Roshie, Angie, Blair, Thomas, Aaron, John"
seed_number <- 
  seed_string |>  
  rlang::hash() |> # hash the string to a hex number
  substr(1, 7) |> # make the number smaller 
  strtoi(base = 16) # convert to decimal

set.seed(seed_number)
```

# Hamilton's Problems

## Exercise 3.1
Is the following MA(2) process covariance-stationary? If so, calculate its autocovariances.
$$Y_t=(1+2.4L+0.8L^2)\varepsilon_t$$
$$\mathbb{E}[\varepsilon_t \varepsilon_{\tau}]=
\begin{cases}
1 & \text{for }t=\tau \\
0 & \text{otherwise}
\end{cases}$$

**Solution:**
We check the conditions for stationarity. First, we take $\varepsilon_t \sim WN(0, 1)$.
$$
\begin{aligned}
\mathbb{E}[Y_t] &= \mathbb{E}[\varepsilon_t+2.4\varepsilon_{t-1}+0.8\varepsilon_{t-2}] \\
&= \mathbb{E}[\varepsilon_t]+2.4\mathbb{E}[\varepsilon_{t-1}]+0.8\mathbb{E}[\varepsilon_{t-2}] \\
&=0 \quad \forall t
\end{aligned}
$$
by virtue of $\varepsilon_t$ being a white noise distribution centered at 0. Next, we check that the variance of this process is finite for all $t$.
$$
\begin{aligned}
\mathbb{V}(Y_t) &= \mathbb{V}(\varepsilon_t+2.4\varepsilon_{t-1}+0.8\varepsilon_{t-2}) \\
&= \mathbb{V}(\varepsilon_t)+2.4^2\mathbb{V}(\varepsilon_{t-1})+0.8^2\mathbb{V}(\varepsilon_{t-2}) \\
&= 1+2.4^2+0.8^2=7.4 < \infty \quad \forall t,
\end{aligned}
$$
where we use the fact that the covariance of $\varepsilon_t$ and any $\varepsilon_{\tau}$, $\tau \neq t$ is 0, and that $\varepsilon_t$ has variance of 1 (for simplicity, it could be any finite $\sigma^2$).

Finally, we check that the covariance between $Y_t$ and some arbitrary $Y_{t-j}$ is dependent only on the size of the gap $j$.
$$
\begin{aligned}
\mathbb{C}\text{ov}(Y_t, Y_{t-j}) &= \mathbb{E}[Y_tY_{t-j}]-\mathbb{E}[Y_t]\mathbb{E}[Y_{t-j}] \\
&= \mathbb{E}[Y_tY_{t-j}]-0.
\end{aligned}
$$

* $j=0$: $\mathbb{C}\text{ov}(Y_t, Y_t)=\mathbb{V}(Y_t)=7.4$ 
* $j=1$: $\mathbb{E}[Y_tY_{t-1}]=\mathbb{E}[(\varepsilon_t+2.4\varepsilon_{t-1}+0.8\varepsilon_{t-2})(\varepsilon_{t-1}+2.4\varepsilon_{t-2}+0.8\varepsilon_{t-3})]$. Using linearity of expectations and the fact that $\mathbb{E}[\varepsilon_t\varepsilon_{\tau}]=0$ for $t \neq \tau$, this simplifies to $2.4\mathbb{E}[\varepsilon_{t-1}^2]+2.4(0.8)\mathbb{E}[\varepsilon_{t-2}^2]=4.32,$ independent of $t$.
* $j=2$: $\mathbb{E}[Y_tY_{t-2}]=\mathbb{E}[(\varepsilon_t+2.4\varepsilon_{t-1}+0.8\varepsilon_{t-2})(\varepsilon_{t-2}+2.4\varepsilon_{t-3}+0.8\varepsilon_{t-4})]$. Again, we use linearity of expectations and see that the only expectation that isn't 0 is the term $\mathbb{E}[\varepsilon_{t-2}^2]$; $\mathbb{E}[Y_tY_{t-2}]=0.8\mathbb{E}[\varepsilon_{t-2}^2]=0.8$, independent of $t$.
* $j > 2$: $\mathbb{C}\text{ov}(Y_t, Y_{t-j}) = 0$ in this case, as there will be no overlapping $\varepsilon$ terms in the expectation of the product calculation. There is only the situation where $\mathbb{E}[\varepsilon_t\varepsilon_{\tau}], t \neq \tau$.
Thus, $\mathbb{C}\text{ov}(Y_t, Y_{t-j})=\gamma_j$, independent of $t$. The MA(2) process $Y_t$ satisfies all three conditions for covariance-stationarity. As an aside, this process is not invertible: the roots of the lag polynomial $P(z)=0.8z^2+2.4z+1$ are $-0.5, -2.5$, which are not both outside of the unit circle. 

The auto-covariances are as described above (again, normalizing $\varepsilon_t$ to have variance of 1:
$$
\mathbb{C}\text{ov}(Y_t, Y_{t-j})=
\begin{cases}
7.4 & \quad \text{if }j=0 \\
4.32 & \quad \text{if }j=1 \\
0.8 & \quad \text{if }j=2\\
0 & \quad \text{else}.
\end{cases}
$$

## Exercise 3.2
Is the following AR(2) process covariance-stationary? If so, calculate its autocovariances.
$$
\begin{aligned}
(1-1.1L+0.18L^2)Y_t &= \varepsilon_t \\
\mathbb{E}[\varepsilon_t \varepsilon_{\tau}]= 
\begin{cases}
1 \quad \text{for }t=\tau \\
0 \quad \text{otherwise}.
\end{cases}
\end{aligned}
$$
**Solution:**

First, we write the process in non-lag operator notation:

$$
\begin{aligned}
(1-1.1L+0.18L^2)Y_t = \varepsilon_t & \Rightarrow Y_t - 1.1LY_t + 0.18L^2Y_t = \varepsilon_t \\ 
& \Rightarrow Y_t - 1.1Y_{t-1} + 0.18Y_{t-2} = \varepsilon_t.
\end{aligned}
$$

Next, we rewrite the process into companion form:

$$
\underbrace{\begin{pmatrix} Y_t \\ Y_{t-1} \end{pmatrix}}_{\xi_t}
=
\underbrace{\begin{pmatrix} 1.1 & -0.18 \\ 1 & 0 \end{pmatrix}}_{F}
\underbrace{\begin{pmatrix} Y_{t-1} \\ Y_{t-2} \end{pmatrix}}_{\xi_{t-1}}
+
\underbrace{\begin{pmatrix} \varepsilon_t \\ 0 \end{pmatrix}}_{\eta_t}.
$$

We find the eigenvalues of $\mathbf{F}$ by solving
$$\begin{aligned}
\det \begin{pmatrix} 1.1 - \lambda & -0.18 \\ 1 &-\lambda\end{pmatrix} &=  0 \\
-1.1 \lambda + \lambda^2 + 0.18 &= 0\\ 
\lambda^2 - 1.1 \lambda + 0.18 &=  0 \\ 
(\lambda - 0.9)(\lambda - 0.2) &=  0.
\end{aligned}$$

Therefore, the two eigenvalues are $\lambda_1 = 0.2$ and $\lambda_2 = 0.9$. Since both eigenvalues have absolute value less than $1$, the process is covariance-stationary.

Next, we calculate the autocovariances. If we follow the approach outlined in the section notes, the algebra becomes quite tedious, so we use the Yule-Walker algorithm instead, which is much cleaner to implement. We start with the AR(2) process and multiple the equation by $Y_t$ and take the expectation:

From $$
\begin{aligned}
Y_t = 1.1Y_{t-1} - 0.18Y_{t-2} + \varepsilon_t & \Rightarrow Y_tY_t = 1.1Y_{t-1}Y_t - 0.18Y_{t-2}Y_t + \varepsilon_tY_t \\ 
& \Rightarrow \underbrace{\mathbb{E}[Y_tY_t]}_{\gamma_0} = 1.1 \underbrace{\mathbb{E}[Y_{t-1}Y_t]}_{\gamma_1} - 0.18\underbrace{\mathbb{E}[Y_{t-2}Y_t]}_{\gamma_2} + \underbrace{\mathbb{E}[\varepsilon_tY_t]}_{=1} \\
& \Rightarrow \gamma_0 = 1.1\gamma_1 - 0.18\gamma_2 + 1,
\end{aligned}
$$ where $\mathbb{E}[\varepsilon_tY_t]$ is equal to 1 because today's shock $\varepsilon_t$ is perfectly correlated with today's value $Y_t$ by a factor of its own variance $\sigma^2 = 1$.

Next, we multiply our process by $Y_{t-1}$:

$$
\begin{aligned}
\begin{aligned}
Y_t = 1.1Y_{t-1} - 0.18Y_{t-2} + \varepsilon_t & \Rightarrow Y_tY_{t-1} = 1.1Y_{t-1}Y_{t-1} - 0.18Y_{t-2}Y_{t-1} + \varepsilon_tY_{t-1} \\ 
& \Rightarrow \underbrace{\mathbb{E}[Y_tY_{t-1}]}_{\gamma_1} = 1.1 \underbrace{\mathbb{E}[Y_{t-1}Y_{t-1}]}_{\gamma_0} - 0.18\underbrace{\mathbb{E}[Y_{t-2}Y_{t-1}]}_{\gamma_1} + \underbrace{\mathbb{E}[\varepsilon_tY_{t-1}]}_{=0} \\
& \Rightarrow \gamma_1 = 1.1\gamma_0 - 0.18\gamma_1,
\end{aligned}
\end{aligned}
$$ where $\mathbb{E}[\varepsilon_tY_{t-1}]$ is equal to 0 because today's shock $\varepsilon_t$ hasn't happened at time $t-1$, so it has zero correlation with $Y_{t-1}$.

Finally, we multiply our process by $Y_{t-2}$:

$$ \begin{aligned} \begin{aligned} Y_t = 1.1Y_{t-1} - 0.18Y_{t-2} + \varepsilon_t & \Rightarrow Y_tY_{t-2} = 1.1Y_{t-1}Y_{t-2} - 0.18Y_{t-2}Y_{t-2} + \varepsilon_tY_{t-2} \\  & \Rightarrow \underbrace{\mathbb{E}[Y_tY_{t-2}]}_{\gamma_2} = 1.1 \underbrace{\mathbb{E}[Y_{t-1}Y_{t-2}]}_{\gamma_1} - 0.18\underbrace{\mathbb{E}[Y_{t-2}Y_{t-2}]}_{\gamma_0} + \underbrace{\mathbb{E}[\varepsilon_tY_{t-2}]}_{=0} \\ & \Rightarrow \gamma_2 = 1.1\gamma_1 - 0.18\gamma_0, \end{aligned} \end{aligned} $$ where $\mathbb{E}[\varepsilon_tY_{t-2}]$ is equal to 0 because today's shock $\varepsilon_t$ hasn't happened at time $t-2$, so it has zero correlation with $Y_{t-1}$.

Now, we have 3 equations and 3 unknowns:

$$
\begin{aligned}
\gamma_0 & = 1.1\gamma_1 - 0.18\gamma_2 + 1 \\ 
\gamma_1 &= 1.1\gamma_0 - 0.18\gamma_1 \\ 
\gamma_2 &= 1.1\gamma_1 - 0.18\gamma_0.
\end{aligned}
$$ The second equation implies $$
1.18\gamma_1 = 1.1\gamma_0
\quad\Rightarrow\quad
\gamma_1 \approx 0.932\,\gamma_0.
$$ The third equation implies

$$
\gamma_2 = 1.1 \gamma_1 - 0.18\gamma_0 \Rightarrow \gamma_2 =  1.1(0.932\gamma_0) - 0.18\gamma_0
\approx 0.845\,\gamma_0.
$$

Substituting $\gamma_1 \approx 0.932\gamma_0$ and $\gamma_2 \approx 0.845\gamma_0$ into the first equation implies $$
\gamma_0 = 1.1(0.932\gamma_0) - 0.18(0.845\gamma_0) + 1
\quad\Rightarrow\quad
\gamma_0 \approx 7.89.
$$

Hence $$
\gamma_1 \approx 0.932(7.89) \approx 7.35,
\qquad
\gamma_2 \approx 0.845(7.89) \approx 6.67.
$$

For $k\geq 3$: $$
\gamma_k = 1.1\gamma_{k-1} - 0.18\gamma_{k-2}.
$$


## Exercise 3.3
A covariance-stationary $AR(p)$ process 
$$
(1-\phi_1L-\phi_2L^2-...-\phi_pL^p)(Y_t-\mu)=\varepsilon_t
$$
has an $MA(\infty)$ representation given by 
$$
\begin{aligned}
(Y_t-\mu) &= \psi(L)\varepsilon_t, \text{ with} \\
\psi(L) &= 1/[1-\phi_1L-\phi_2L^2-...-\phi_pL^p] \text{ or} \\
[1-\phi_1L-\phi_2L^2-...-\phi_L^p]&[\psi_0+\psi_1L+\psi_2L^2+...]=1.
\end{aligned}
$$
In order for this equation to be true, the implied coefficient on $L^0$ must be unit and the coefficients on $L^1, L^2, ...$ must be zero. Write out these conditions explicitly and show that they imply a recursive algorithm for generating the $MA(\infty)$ weights $\psi_0, \psi_1, ...$. Show that this recursion is algebraically equivalent to setting $\psi_j$ equal to the $(1,1)$ element of the matrix \textbf{F} raised to the $j$th power.

**Solution:**

If we expand the product and make explicit the conditions for the $MA(\infty)$ representation to hold true, we obtain a series of equations on the respective lag terms:
$$
\begin{aligned}
L^0&: \psi_0=1 \\
L^1&: \psi_1-\psi_0\phi_1 = 0 \Rightarrow \psi_1 = \phi_1 \\
L^2&: \psi_2-\phi_1\psi_1-\phi_2\psi_0 = 0 \Rightarrow \psi_2 = \phi_2 + \phi_1^2 \\
... \\
L^j&: \psi_j = \phi_1 \psi_{j-1}+\phi_2\psi_{j-2}+...+\phi_p\psi_{j-p} \\
& \psi_j= \sum_{k=1}^p \phi_k \psi_{j-k}.
\end{aligned}
$$
As in lecture, we can define the \textbf{F} matrix as
$$
\begin{bmatrix}
\phi_1 & \phi_2 & ... & \phi_{p-1} & \phi_p \\
1 & 0 & ... & 0 & 0 \\
0 & 1 & ... & 0 & 0 \\
\vdots & \vdots & \vdots & \vdots & \vdots \\
0 & 0 & ... & 1 & 0,
\end{bmatrix}
$$
where the first row is $\phi_1$ through $\phi_p$, the other elements of the final column are 0's, and the remainder of the matrix is $I_{p-1}$ (the identity).

Then, $F_{11}=\phi_1=\psi_1$, by construction. The (1,1) element of $F^2$ is $\phi_1^2+\phi_2=\phi_1\psi_1+\phi_2=\psi_2$, as desired. The (1,1) element of $F^p$ is $\phi_1\psi_{p-1}+\phi_2\psi_{p-2}+...+\phi_p\psi_0$, which from the $MA(\infty)$ conditions we have shown to be $\psi_p$. In particular, the first two elements of the first row of the product $F^{j-1}$ are constructed as follows: $[\psi_{j-1} , \phi_2\psi_{j-2}+\phi_3\psi_{j-3} , ...]$, which when taking the dot product with $F$ leaves $\sum_{k=1}^p \phi_k\psi_{j-k}=\psi_j$ as the (1,1) element. Thus, the recursion algorithm is equivalent to setting $\psi_j$ to be the (1,1) element of $F^j$.

## Exercise 11.5
Consider the following bivariate VAR:
$$
\begin{aligned}
y_{1,t} &= 0.3y_{1,t-1}+0.8y_{2,t-1}+\varepsilon_{1,t} \\
y_{2,t}&= 0.9y_{1,t-1}+0.4y_{2,t-1}+\varepsilon_{2,t}
\end{aligned}
$$
with $\mathbb{E}[\varepsilon_{1,t}\varepsilon_{1,\tau}]=1$ for $t=\tau$ and 0 otherwise, $\mathbb{E}[\varepsilon_{2,t}\varepsilon_{2,\tau}]=2$ for $t=\tau$ and 0 otherwise, and $\mathbb{E}[\varepsilon_{1,t}\varepsilon_{2,\tau}]=0$ for all $t,\tau$.

(a) Is this system covariance-stationary?
(b) Calculate $\mathbf{\Psi}_s=\frac{\partial y_{t+s}}{\partial\varepsilon_t'}$ for $s=0,1,2$. What is the limit as $s \rightarrow \infty$?
(c) Calculate the fraction of the MSe of the two-period-ahead forecast error for variable 1,
$$
\mathbb{E}[y_{1,t+2}-\mathbb{E}[y_{1,t+2}|y_t,y_{t-1},...]]^2,
$$
that is due to $\varepsilon_{1,t+1}$ and $\varepsilon_{1,t+2}$.

**Solution:**

(a) This system is not covariance-stationary. We can rewrite it in vector form:
$$
\begin{bmatrix}
Y_{1,t} \\ Y_{2,t} 
\end{bmatrix}
= 
\begin{bmatrix}
0.3 & 0.8 \\ 0.9 & 0.4
\end{bmatrix}
\begin{bmatrix}
Y_{1,t-1} \\
Y_{2,t-1}
\end{bmatrix} +
\begin{bmatrix}
\varepsilon_{1,t} \\
\varepsilon_{2,t}
\end{bmatrix}.
$$
The eigenvalues of the $\Pi$ matrix (coefficient on $Y_{t-1}$) are found by setting the determinant of $\Pi-\lambda I$ to be 0; we get $(0.3-\lambda)(0.4-\lambda)-0.8*0.9$, yielding the eigenvalues of 1.2 and -0.5. One eigenvalue of $\pi$ (the companion form) is outside of the unit circle, meaning that this bivariate VAR process is not covariance-stationary.

(b) We consider $\Psi_s=\frac{\partial y_{t+s}}{\partial \varepsilon_t'}$ for various values of $s$.

* When $s=0$: $\Psi_0=\frac{\partial y_{t}}{\partial \varepsilon_t'}$.

$$
\begin{aligned}
\frac{\partial y_{1,t}}{\partial \varepsilon_{1,t}} &= 1 \quad \frac{\partial y_{1,t}}{\partial \varepsilon_{2,t}} &= 0\\
\frac{\partial y_{2,t}}{\partial \varepsilon_{1,t}} &= 0 \quad \frac{\partial y_{2,t}}{\partial \varepsilon_{2,t}} &= 1 \\
\Psi_0 &= 
\begin{bmatrix}
1 & 0 \\ 0 & 1 
\end{bmatrix}
\end{aligned}
$$

* When $s=1$: $\Psi_1=\frac{\partial y_{t+1}}{\partial \varepsilon_t'}$.

$$
\begin{aligned}
y_{t+1} &= \Pi y_t + \varepsilon_{t+1} \\
y_t &= \Pi y_{t-1} + \varepsilon_t \Rightarrow y_{t+1} = \Pi[\Pi y_{t-1}+\varepsilon_t]+\varepsilon_{t+1} \\
\frac{\partial y_{t+1}}{\partial \varepsilon_t'} &= \Pi = \begin{bmatrix} 0.3 & 0.8 \\ 0.9 & 0.4 \end{bmatrix}
\end{aligned}
$$
* When $s=2$: $\Psi_2=\frac{\partial y_{t+2}}{\partial \varepsilon_t'}$. We can apply a similar recursion to find the proper coefficient on on $\varepsilon_t$ and show that

$$
\begin{aligned}
y_{t+1} &= \Pi[\Pi[\Pi y_{t-1}+\varepsilon_{t}]+\varepsilon_{t+1}]+\varepsilon_{t+2} \\
\frac{\partial y_{t+2}}{\partial \varepsilon_t'} &= \Pi^2=\begin{bmatrix}0.81 & 0.56 \\ 0.63 & 0.88 \end{bmatrix}
\end{aligned}
$$
* When $s \rightarrow \infty$, the recursive pattern says that $\Psi_s = \lim_{s \rightarrow \infty} \Pi^s$. We can simplify the term using the spectral decomposition, as $\Pi$ has a full set of linearly independent eigenvectors. Below, $X$ is the matrix whose columns are the eigenvectors of $\Pi$, and $\Lambda$ is the diagonal matrix of $\Pi$'s eigenvalues. We can immediately see that this term will diverge as $s \rightarrow \infty$, as $1.2 > 1$---shocks grow exponentially rather than dying out. So, this bivariate VAR process is explosive *and* not covariance-stationary

$$
\begin{aligned}
\Pi^s &= X \Lambda^s X^{-1}
\end{aligned}
$$

(c) To determine the two-period ahead forecast error for $y_1$, we can first recursively substitute to simplify the inner expectation term

$$
\begin{aligned}
\mathbb{E}_t[y_{1,t+2}|y_t, y_{t-1},...] &= \mathbb{E}_t[0.3y_{1,t+1}+0.8y_{2,t+1}+\varepsilon_{1,t+2}|y_t, y_{t-1},...] \\
&= \mathbb{E}[0.3(0.3y_{1,t}+0.8y_{2,t}+\varepsilon_{1, t+1})+0.8(0.9y_{1,t}+0.4y_{2,t}+\varepsilon_{2,t+1})+\varepsilon_{1,t+2}|y_t,y_{t-1},...] \\
&= (0.09+0.72) \mathbb{E}_t[y_{1,t}|\cdot]+(0.24+0.32)\mathbb{E}_t[y_{2,t}|\cdot] \\
& \quad + 0.3\mathbb{E}_t[\varepsilon_{1,t+1}|\cdot] +0.8\mathbb{E}_t[\varepsilon_{2,t+1}|\cdot] +\mathbb{E}_t[\varepsilon_{1,t+2}|\cdot]\\
&= 0.81 y_{1,t} + 0.56 y_{2,t}.
\end{aligned}
$$

Now, plugging this into the MSE expression---there seems to be a typo, where the MSE should be $\mathbb{E}[(x-\mathbb{E}[x])^2]$, not $(\mathbb{E}[x-\mathbb{E}[x]])^2=0$---and using the $y_{1,t+2}$ recursion derived above,

$$
\begin{aligned}
\mathbb{E}_t[(y_{1,t+2}-\mathbb{E}_t[y_{1,t+2}|y_t, y_{t-1},...])^2] &= \mathbb{E}_t[(y_{1,t+2}-(0.81y_{1,t}+0.56 y_{2,t}))^2] \\
&= \mathbb{E}_t[(0.81y_{1,t}+0.56y_{2,t}+0.3\varepsilon_{1,t+1}+0.8\varepsilon_{2,t+1}+\varepsilon_{1,t+2}\\
& \quad \quad - 0.81y_{1,t}-0.56 y_{2,t})^2] \\
&= \mathbb{E}_t[(0.3\varepsilon_{1,t+1}+0.8\varepsilon_{2, t+1}+\varepsilon_{1,t+2})^2]\\
& \quad \text{as } \mathbb{E}_t[\varepsilon_{i,t}]=0, \text{ cancelling out the }\mathbb{E}[\cdot]^2 \text{ term of variance}  \\
&= \mathbb{V}(0.3\varepsilon_{1,t+1}+0.8\varepsilon_{2,t+1}+\varepsilon_{1,t+2}) \\
& \quad \text{by independence of } \varepsilon_{i,t} \text{ terms} \\
&= 0.3^2\mathbb{V}(\varepsilon_{1,t+1})+0.8^2\mathbb{V}(\varepsilon_{2, t+1})+\mathbb{V}(\varepsilon_{1,t+2}) \\
&= 0.09(1)+0.64(2)+1=2.37.
\end{aligned}
$$
Then, the share of the mean-squared error coming from $\varepsilon_{1,t+1}$ is $0.09/2.37 \approx 0.038$, while the share from $\varepsilon_{1,t+2}$ is $1/2.37 \approx 0.422$.

# The Permanent Income Model

## 1. Finding the Autocovariances

Conditional on the information available at time $t$, all of the future random 
variables are mean zero, so the agent's 
expected income in any period $j\geq 0$ will their current permanent component of
income, $x_t$. The transitory shock includes an $\epsilon_t$ that moves current
period income from its long run average. So $E_t[y_{t+j+1}]=x_t$. The time $t+1$
expectations are themselves a random variable from the perspective of period $t$
because they depend on the news revealed in period $t+1$. In particular, in period
$t+1$, a new value $x_{t+1}$ will be revealed, recentering the agents expectations
of all future periods. If $j=0$, then from period $t+1$'s perspective, $y_{t+1}$
  is observed with certainty. But all future $j\geq 1$ are centered at $x_{t+1}$.

Noting that $x_{t+1}=x_t+\eta_{t+1}$ and $y_{t+1}=x_{t}+\eta_{t+1}+\varepsilon_{t+1}$, we see that:
  
  $$
  E_{t+1}[y_{t+j+1}]-E_t[y_{t+j+1}]=
  \begin{cases}
x_{t}+\eta_{t+1}+\varepsilon_{t+1} - x_t & j=0 \\
x_t+\eta_{t+1}-x_t & j \geq 1
\end{cases}
$$
  
  Which obviously simplifies to:
  
  $$
  E_{t+1}[y_{t+j+1}]-E_t[y_{t+j+1}]=\begin{cases}
\eta_{t+1}+\varepsilon_{t+1}& j=0 \\
\eta_{t+1} & j \geq 1
\end{cases}
$$
  With that in mind, we can conveniently express $\Delta c_{t+1}\equiv c_{t+1}-c_{t}$ as:
  
  $$
  \begin{aligned}
\Delta c_{t+1}&=(1-\beta)\sum_{j=0}^\infty\beta^j\left( E_{t+1}[y_{t+j+1}]-E_t[y_{t+j+1}]\right)\\
&=(1-\beta)\beta^0\left( E_{t+1}[y_{t+0+1}]-E_t[y_{t+0+1}]\right)+(1-\beta)\sum_{j=1}^\infty\beta^j\left( E_{t+1}[y_{t+j+1}]-E_t[y_{t+j+1}]\right)\\
&=(1-\beta)\left( \eta_{t+1}+\varepsilon_{t+1}\right)+(1-\beta)\sum_{j=1}^\infty\beta^j\eta_{t+1}\\
&=(1-\beta)\left(\varepsilon_{t+1}\right)+(1-\beta)\eta_{t+1}+(1-\beta)\sum_{j=1}^\infty\beta^j\eta_{t+1} \\
&=(1-\beta)\left(\varepsilon_{t+1}\right)+(1-\beta)\sum_{j=0}^\infty\beta^j\eta_{t+1} \\
&=(1-\beta)\left(\varepsilon_{t+1}\right)+\eta_{t+1}(1-\beta)\sum_{j=0}^\infty\beta^j \\
&=(1-\beta)\left(\varepsilon_{t+1}\right)+\eta_{t+1}(1-\beta)\frac{1}{1-\beta} \\
&=(1-\beta)\varepsilon_{t+1}+\eta_{t+1} \\
\end{aligned}
$$
  
  If we define $u_{t}=(1-\beta)\varepsilon_{t+1}+\eta_{t+1}\sim\mathcal{N}(0,(1-\beta)^2\sigma_\varepsilon^2+\sigma_\eta^2)$, then we can think of $\Delta c_{t}$ as white noise process with mean 0 and variance: $(1-\beta)^2\sigma_\varepsilon^2+\sigma_\eta^2)$.

From this, it is clear to see that the IRFs and autocovariances for all $j\neq 0$
  will be zero. There is absolutely no information passed between periods (at least
                                                                           for the per-period change in consumption).

If $j=0$, then we have already expressed the variance as $(1-\beta)^2\sigma_\varepsilon^2+\sigma_\eta^2)$ Clearly, these are increasing in the variances of the primitives. But interesting as the agent
becomes more patient ($\beta\to 1$), the variance in consumption reduces to only variance
in the permanent component (because people no longer respond to the transitory shocks.)

And the IRFs are simply the partials with respect to $\varepsilon$ which is $(1-\beta)$
  and for $\eta$ which is $1$.

Assuming $\beta\in (0,1)$ (people strictly discount the future), then the magnitude
of a transitory shock on the change in consumption is less than the same sized shock
to permanent income. The permanent IRFs are invariant to model parameters, whereas
as individuals become more patient $\beta\to 1$, the IRF for the transitory shock
approaches zero (little change to previous period consumption).

## 2. The Vector Process

Noting that 

$$
  \begin{aligned}
\Delta y_{t+1}&=(x_{t+1}+\varepsilon_{t+1})-(x_{t}+\varepsilon_{t}) \\
&=(x_t+\eta_{t+1}+\varepsilon_{t+1})-(x_{t}+\varepsilon_{t})\\
&=\eta_{t+1}+\varepsilon_{t+1}-\varepsilon_{t}%\\
%&=\eta_{t+1}+\varepsilon_{t+1}-\varepsilon_{t}+\beta\varepsilon_{t+1}-\beta\varepsilon_{t+1}\\
%&=\eta_{t+1}+\varepsilon_{t+1}-\varepsilon_{t}+\beta\varepsilon_{t+1}-\beta\varepsilon_{t+1}\\
%&=\underbrace{\eta_{t+1}+(1-\beta)\varepsilon_{t+1}}_{\Delta %c_{t+1}}-\varepsilon_{t}+\beta\varepsilon_{t+1}\\
\end{aligned}
$$
  
  Decrementing the indices by one period, we can see that:
  
  $$
  \begin{bmatrix}\Delta c_{t}\\ \Delta y_{t}\end{bmatrix}=\begin{bmatrix} (1-\beta)\varepsilon_{t}+\eta_{t} \\ \eta_{t}+\varepsilon_{t}-\varepsilon_{t-1} \end{bmatrix}
  $$
    
    From this form, it would be natural to represent using matrix notation:
    
    $$
    \mathbf{\Delta}_t = \mathbf{\Psi}_0\mathbf{u}_t+ \mathbf{\Psi}_1\mathbf{u}_{t-1}
    $$
      
      Where
    $$
      \mathbf{\Psi}_0=\begin{bmatrix}(1-\beta) &1\\ 1 & 1 \end{bmatrix}, \mathbf{\Psi}_1=\begin{bmatrix}0 & 0 \\
      -1 & 0 \\ \end{bmatrix},\mathbf{u}_t\sim WN\left(\mathbf{0},\begin{bmatrix}\sigma_\varepsilon^2 & 0\\ 0 & \sigma_\eta^2\end{bmatrix}\right)
      $$
        
        Which is a rather simple vector moving average equation, but it still suffers
      from an identification problem because it has 8 parameters including some on the
      contemporaneous shocks. One way to take care that is to redefine the contemporaneous
      shocks so that they're no longer orthogonal but there is no longer a $\mathbf{\Psi}_0$ matrix.

Defining $\mathbf{e}_t\equiv\mathbf{\Psi}_0 \mathbf{u}_t$, linearity of the normal family
shows that:

$$
\begin{aligned}
\mathbf{e}_t&\sim WN\left(\mathbf{0},\begin{bmatrix}(1-\beta) &1\\ 1 & 1 \end{bmatrix}\begin{bmatrix}\sigma_\varepsilon^2 & 0\\ 0 & \sigma_\eta^2\end{bmatrix}\begin{bmatrix}(1-\beta) &1\\ 1 & 1 \end{bmatrix}'\right) \\ 
&\sim WN\left(\mathbf{0},\begin{bmatrix}(1-\beta)\sigma_\varepsilon^2 &\sigma_\eta^2\\ \sigma_\varepsilon^2 & \sigma_\eta^2 \end{bmatrix}\begin{bmatrix}(1-\beta) &1\\ 1 & 1 \end{bmatrix}\right)\\
&\sim WN\left(\mathbf{0},\begin{bmatrix}(1-\beta)^2\sigma_\varepsilon^2+\sigma_\eta^2 &(1-\beta)\sigma_\varepsilon^2+\sigma_\eta^2\\ (1-\beta)\sigma_\varepsilon^2+\sigma_\eta^2 & \sigma_\varepsilon^2+\sigma_\eta^2 \end{bmatrix}\right)
\end{aligned}
$$
  
  and that we solve for last period's shocks:

$$
\mathbf{e}_{t-1}=\mathbf{\Psi}_0\mathbf{u}_{t-1}\implies \mathbf{u}_{t-1}=\mathbf{\Psi}_0^{-1}\mathbf{e}_{t-1}
$$

With that we can re-express our process as:

$$
\mathbf{\Delta}_t = \mathbf{e}_t+ \mathbf{\Psi}_1\mathbf{\Psi}_0^{-1}\mathbf{e}_{t-1}
$$

Noting that using the familiar $2\times 2$ inverse identity, we see that:

$$
\begin{aligned}
\mathbf{\Psi}_0^{-1}&=\frac{1}{\det(\mathbf{\Psi}_0)}\begin{bmatrix}1 & -1\\ -1 & (1-\beta)\end{bmatrix} \\
&=\frac{1}{(1-\beta)\cdot1-1\cdot1}\begin{bmatrix}1 & -1\\ -1 & (1-\beta)\end{bmatrix}\\
&=\frac{1}{-\beta}\begin{bmatrix}1 & -1\\ -1 & (1-\beta)\end{bmatrix}\\
\end{aligned}
$$

And combining $\mathbf{\Psi}_1\mathbf{\Psi}_0^{-1}$ into a single matrix $\tilde{\mathbf{\Psi}}_1$, we see that:

$$
\begin{aligned}
\tilde{\mathbf{\Psi}}_1&=\mathbf{\Psi}_1\mathbf{\Psi}_0^{-1}\\
&=\begin{bmatrix}0 & 0 \\
-1 & 0 \\ \end{bmatrix}\frac{1}{-\beta}\begin{bmatrix}1 & -1\\ -1 & (1-\beta)\end{bmatrix}\\
&=\frac{1}{-\beta}\begin{bmatrix}0 & 0 \\
-1 & 0 \\ \end{bmatrix}\begin{bmatrix}1 & -1\\ -1 & (1-\beta)\end{bmatrix}\\
&=\frac{1}{-\beta}\begin{bmatrix}0 & 0 \\
-1 & 1 \\ \end{bmatrix}\\
&=\begin{bmatrix}0 & 0 \\
\frac{1}{\beta} & \frac{1}{-\beta} \\ \end{bmatrix}\\
\end{aligned}
$$

So the final form for the vector moving average process is:

$$
\begin{gathered}
\mathbf{\Delta}_t = \mathbf{e}_t+ \tilde{\mathbf{\Psi}}_1\mathbf{e}_{t-1}\\
s.t. \\
\tilde{\mathbf{\Psi}}_1=\begin{bmatrix}0 & 0 \\
\frac{1}{\beta} & \frac{1}{-\beta} \\ \end{bmatrix}, \mathbf{e}_t\sim WN\left(\mathbf{0},\begin{bmatrix}(1-\beta)^2\sigma_\varepsilon^2+\sigma_\eta^2 &(1-\beta)\sigma_\varepsilon^2+\sigma_\eta^2\\ (1-\beta)\sigma_\varepsilon^2+\sigma_\eta^2 & \sigma_\varepsilon^2+\sigma_\eta^2 \end{bmatrix}\right)
\end{gathered}
$$



## 3. Simulation Exercise

Below I simulate 10,000 steps of the simulation using the VMA(1) form discussed
in the previous question (and correctly handling the correlated random draws).

```{r}
capT = 10e3
beta = .99
sigma_e = .05
sigma_h = .05

# Covariance structure of the VMA representation
# this is not the standard deviations
Sigma = matrix(c((1-beta)^2*sigma_e^2 + sigma_h^2, (1-beta)*sigma_e^2 + sigma_h^2,
                 (1-beta)*sigma_e^2 + sigma_h^2, sigma_e^2 + sigma_h^2), 
               byrow = T, nrow = 2)

Psy_tilde = matrix(c(0,0,
               1/beta,-1/beta), byrow = T, nrow = 2)
# generate shocks
e_draws <- mvtnorm::rmvnorm(capT, mean = c(0,0), sigma = Sigma)

# initialize matrices for output
v_mat <- matrix(nrow = nrow(e_draws), ncol = 2)
colnames(v_mat) <- c("Delta c","Delta y")

for (i in 1:capT){
  if (i == 1){
    # if first observation, treat last period's shocks as 0
e_tm1 <- c(e = 0, h = 0) |> matrix(ncol = 1)
}else{
  # otherwise just get shocks from last period
  e_tm1 <- e_draws[i-1,] |> matrix(ncol = 1)
}

# get current shocks
e_t <- e_draws[i,] |> matrix(ncol = 1)

v = e_t + Psy_tilde%*%e_tm1

v_mat[i,] <- t(v)
}

library(ggplot2)
ggplot()+
  geom_point(aes(x=v_mat[,1], y = v_mat[,2], col = 1:capT), alpha = .2)+
  labs(color = "Time",
       x = "Delta c",
       y = "Delta y")

```

Now I'll compute a misspecified VAR(5) model and compare its IRFs to compare them
to the true values. To quickly recap, note that our structural VAR equations were
of the form

$$
\mathbf{\Delta}_t = \mathbf{\Psi}_0\mathbf{u}_t+ \mathbf{\Psi}_1\mathbf{u}_{t-1}
$$

Where
$$
\mathbf{\Psi}_0=\begin{bmatrix}(1-\beta) &1\\ 1 & 1 \end{bmatrix}, \mathbf{\Psi}_1=\begin{bmatrix}0 & 0 \\
-1 & 0 \\ \end{bmatrix},\mathbf{u}_t\sim WN\left(\mathbf{0},\begin{bmatrix}\sigma_\varepsilon^2 & 0\\ 0 & \sigma_\eta^2\end{bmatrix}\right)
$$

So the true (causal) impulse responses would be given by:

$$
\frac{\partial\mathbf{\Delta}_{t+0}}{\partial \mathbf{u}}=\begin{bmatrix}(1-\beta) &1\\ 1 & 1 \end{bmatrix}, 
\frac{\partial\mathbf{\Delta}_{t+1}}{\partial \mathbf{u}}=\begin{bmatrix}0 &0\\ -1 & 1 \end{bmatrix},
\frac{\partial\mathbf{\Delta}_{t+j}}{\partial \mathbf{u}}=\begin{bmatrix}0 &0\\ 0 & 0 \end{bmatrix},
$$
For all $j\geq 2$. 

So of interest to us is the IRFs for $\Delta c_{t}$, so those should be in terms
of $\Delta c_{t}$, $(1-\beta), 1,$ then zero afterwards. And then for $y$, $1$ in
the first period, then zero afterwards.

After implementing the VAR, I'll take a shortcut and use the `vars` package's
off-the-shelf IRF function. And compare it to the patterns above

```{r unrestricted-VAR}
# estimate a var with no trend component
my_p = 5
m2=vars::VAR(v_mat, p = my_p, type = "none")

m2_irf <- vars::irf(m2, ortho = F)$irf$Delta.c
true_irf_c <- c(1-beta, 1, rep(0, nrow(m2_irf)-2))
true_irf_y <- c(1, rep(0, nrow(m2_irf)-1))
```

```{r plot-irfs, echo = F}
ggplot()+
  geom_line(aes(x=0:(nrow(m2_irf)-1), y=true_irf_c, linetype = "True IRF", col = "Transitory"))+
  geom_line(aes(x=0:(nrow(m2_irf)-1), y=m2_irf[,'Delta.c'], linetype = "Estimated IRF", col = "Transitory"))+
  geom_line(aes(x=0:(nrow(m2_irf)-1), y=true_irf_y, linetype = "True IRF", col = "Permanent"))+
  geom_line(aes(x=0:(nrow(m2_irf)-1), y=m2_irf[,'Delta.y'], linetype = "Estimated IRF", col = "Permanent")) +
  ggtitle("Impulse Response of Consumption to Transitory & Permanent Shocks")+
  xlab("Time Since Shock")+
  ylab("Change in Consumption")
```

It's actually relatively close to the true IRFs. We also ran this with $p=50$,
and then the IRFs are indistinguishable. To get even closer we could
use Cholesky decomposition giving precedent to the $\varepsilon$ shocks since
they are completely exogenous.

# The Kalman Filter

## 1. Analytical Moments

We are given the ARMA(1,1) process:

$$
y_t = \phi y_{t-1} + \varepsilon_t+\gamma \varepsilon_{t-1}
$$

In lag operator notation, we have:

$$
\begin{aligned}
y_t &= \phi L y_{t} + \varepsilon_t+\gamma L \varepsilon_{t}\\
(1-\phi L)y_t &= (1+\gamma L) \varepsilon_{t}
\end{aligned}
$$

The multiplicative inverse of $(1-\phi L)$ is $(1+\phi L+\phi^2 L^2+\ldots)$, and 
so we can alternately express the ARMA process as an MA($\infty$):

$$
\begin{aligned}
(1+\phi L+\phi^2 L^2-\ldots)(1-\phi L)y_t &= (1+\phi L+\phi^2 L^2+\ldots)(1+\gamma L) \varepsilon_{t}\\
y_t&=(1+\phi L+\phi^2 L^2+\ldots)(1+\gamma L) \varepsilon_{t}\\
&=[(1+\phi L+\phi^2 L^2+\phi^3L^3\ldots) + (\gamma L+\gamma\phi L^2+\gamma\phi^2 L^3+\ldots)]\varepsilon_{t} \\
&=[\underbrace{1}_{\psi_0}+\underbrace{\phi^0(\phi+\gamma)}_{\psi_1}L+\underbrace{\phi^1(\phi+\gamma)}_{\psi_2}L^2+\underbrace{\phi^2(\phi+\gamma)}_{\psi_3}L^3+\ldots]\varepsilon_{t}\\
&=\psi(L)\varepsilon_t
\end{aligned}
$$

Another equivalent expression is
$$
y_t = \psi(L)\varepsilon_t = \varepsilon_{t} + (\phi + \gamma_1) \sum_{j=1}^\infty \phi^{j-1} \varepsilon_{t-j}
$$

Assuming at every point that each step is well defined.

In this form, it is much easier to compute the various moments of the distribution.

Namely we have:

$$
\begin{aligned}
E[y_t]&=E\left[[1+\phi^0(\phi+\gamma)L+\phi^1(\phi+\gamma)L^2+\phi^2(\phi+\gamma)L^3-\ldots]\varepsilon_{t}\right]\\
&=[1+\phi^0(\phi+\gamma)L+\phi^1(\phi+\gamma)L^2+\phi^2(\phi+\gamma)L^3-\ldots]E[\varepsilon_{t}]\\
&=[1+\phi^0(\phi+\gamma)L+\phi^1(\phi+\gamma)L^2+\phi^2(\phi+\gamma)L^3-\ldots]0\\
&= 0
\end{aligned}
$$
From above, define $\psi_j$ to be
$$\psi_j = \begin{cases}
1 &\text{if } j = 0\\
(\phi + \gamma) \phi^{j-1} &\text{if } j > 0

\end{cases}$$
The variance of any $MA(\infty)$ process is given by:

$$
\begin{aligned}
\gamma_0&=\sigma^2\sum_{j=0}^\infty \psi_j^2 \\
&=\sigma^2\left(\psi_0^2+\sum_{j=1}^\infty \psi_j^2\right) \\
&=\sigma^2\left(1^2+\sum_{j=1}^\infty (\phi^{j-1}(\phi+\gamma))^2\right)\\
&=\sigma^2\left(1+(\phi+\gamma))^2\sum_{j=1}^\infty {\phi^{j-1}}^2\right)\\
&=\sigma^2\left(1+(\phi+\gamma))^2\sum_{j=0}^\infty {\phi^j}^2\right)\\
&=\sigma^2\left(1+(\phi+\gamma))^2\sum_{j=0}^\infty {\phi^2}^j\right)\\
&=\boxed{\sigma^2\left(1+\frac{(\phi+\gamma)^2}{1-\phi^2}\right)}\\
\end{aligned}
$$

requiring that $\phi^2<1\implies |\phi|<1$ for convergence of the geometric series.

The $j$th autocovariance of the MA($\infty$) is given by:

$$
\begin{aligned}
\gamma_j &= \sigma^2\sum_{k=0}^\infty\psi_k\psi_{j+k} \\
&=\sigma^2\left(\psi_0\psi_j + \sum_{k=1}^\infty \psi_k\psi_{j+k} \right) \\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +\sum_{k=1}^\infty \phi^{k-1}(\phi+\gamma)\phi^{j+k-1}(\phi+\gamma)\right) \\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +\sum_{\kappa=0}^\infty \phi^{\kappa}(\phi+\gamma)\phi^{j+\kappa}(\phi+\gamma)\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\sum_{\kappa=0}^\infty \phi^{\kappa}\phi^{j+\kappa}\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\sum_{\kappa=0}^\infty \phi^{2\kappa}\phi^{j}\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\phi^{j}\sum_{\kappa=0}^\infty \phi^{2\kappa}\right)\\
&=\sigma^2\left(1\cdot \phi^{j-1}(\phi+\gamma) +(\phi+\gamma)^2\phi^{j}\frac{1}{1-\phi^2}\right)\\
&=\sigma^2\phi^{j-1}(\phi+\gamma) \left(1 +(\phi+\gamma)\phi\frac{1}{1-\phi^2}\right)\\
&=\sigma^2\phi^{j-1}(\phi+\gamma) \left(1 +(\phi+\gamma)\frac{\phi}{1-\phi^2}\right)\\
&=\sigma^2\phi^{j-1}(\phi+\gamma) \left(\frac{{1-\phi^2}}{{1-\phi^2}} +\frac{\phi^2+\phi\gamma}{1-\phi^2}\right)\\
&=\boxed{\sigma^2\phi^{j-1}(\phi+\gamma) \left(\frac{{1+\phi\gamma}}{{1-\phi^2}}\right)}
\end{aligned}
$$

Truly a bizarre expression that leaves little room for intuition. However, the series
convergence there also only required that $\phi^2<1$, so $|\phi|<1$ is still sufficient to
ensure covariance stationarity.

## 2. Conditional Expectations

Note that if we could observe $y_0$ and $\varepsilon_0$, we can compute $\varepsilon_1$ exactly using
the equation of motion with:

$$
\varepsilon_1 = y_1-\phi y_0-\gamma \varepsilon_0
$$

Then we could proceed inductively to compute all future $\varepsilon_\tau$ for
$\tau\in \{1,\ldots,t\}$ using:

$$
\varepsilon_\tau = y_\tau-\phi y_{\tau-1}-\gamma \varepsilon_{\tau -1}
$$

So given the observed data $\mathcal{I}_t$, and some guesses about
$\varepsilon_0$ and $y_0$, we can compute the exact history of $\varepsilon$s. We
do not need to condition on the $\varepsilon$s, so long as we condition on the other three.

Now consider the first time step after $t$, time $t+1$. Conditional on $\mathcal{I}_t$, 
$\varepsilon_0$, and $y_0$ we can think of the conditional expectation of $y_{t+1}$.

$$
\begin{aligned}
E[y_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]&=E[\phi y_t+\varepsilon_{t+1}+\gamma\varepsilon_t|\mathcal{I}_t,y_0,\varepsilon_0]\\
&=E[\phi y_t|\mathcal{I}_t,y_0,\varepsilon_0]+E[\varepsilon_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]+E[\gamma\varepsilon_t|\mathcal{I}_t,y_0,\varepsilon_0]\\
&=\phi y_t+E[\varepsilon_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]+\gamma\varepsilon_t\\
&=\phi y_t+0+\gamma\varepsilon_t\\
&=\phi y_t+\gamma\varepsilon_t\\
\end{aligned}
$$

Where the middle expectation can be replaced with a 0 because $\varepsilon_{t+1}$
is white noise.

And then for the second step:

$$
\begin{aligned}
E[y_{t+2}|\mathcal{I}_t,y_0,\varepsilon_0]&=E[\phi y_{t+1}+\varepsilon_{t+2}+\gamma\varepsilon_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]\\
&=E[\phi y_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]+E[\varepsilon_{t+2}|\mathcal{I}_t,y_0,\varepsilon_0]+E[\gamma\varepsilon_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]\\
&=E[\phi y_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]+0+0\\
&=\phi E[y_{t+1}|\mathcal{I}_t,y_0,\varepsilon_0]+0+0\\
&=\phi\left(\phi y_t+\gamma\varepsilon_t\right)
\end{aligned}
$$

And for any stage $k$ steps after $t$ we see that:

$$
\begin{aligned}
E[y_{t+k}|\mathcal{I}_t,y_0,\varepsilon_0]&=E[\phi y_{t+(k-1)}+\varepsilon_{t+k}+\gamma\varepsilon_{t+(k-1)}|\mathcal{I}_t,y_0,\varepsilon_0]\\
&=E[\phi y_{t+(k-1)}|\mathcal{I}_t,y_0,\varepsilon_0]+E[\varepsilon_{t+k}|\mathcal{I}_t,y_0,\varepsilon_0]+E[\gamma\varepsilon_{t+(k-1)}|\mathcal{I}_t,y_0,\varepsilon_0]\\
&=E[\phi y_{t+(k-1)}|\mathcal{I}_t,y_0,\varepsilon_0]+0+0\\
&=\boxed{\phi^{k-1}\left(\phi y_t+\gamma\varepsilon_t\right)}
\end{aligned}
$$

## 3. Conditional Likelihood

The natural way to approach this problem is to use the definition of the likelihood
and the chain rule of probability

From that we can see:

$$
\begin{aligned}
p(\mathcal{I}_t|y_0,\varepsilon_0)&=p(y_t,\ldots,y_1|y_0,\varepsilon_0)\\
&=p(y_t,|y_0,\varepsilon_0,y_{t-1},\ldots,y_1)p(y_{t-1},\ldots,y_1|y_0,\varepsilon_0)\\
&=p(y_t,|y_0,\varepsilon_0,y_{t-1},\ldots,y_1)p(y_{t-1}|y_0,\varepsilon_0,y_{t-2},\ldots,y_1)p(y_{t-2},\ldots,y_1|y_0,\varepsilon_0)\\
\vdots & \\
&=\prod_{\tau=1}^tp(y_\tau|y_0,\varepsilon_0,y_{\tau-1},\ldots)
\end{aligned}
$$

Looking at any particular term in this product of densities, we can see it is
the same as predicting the density of the next period, $\tau$ conditional on the
history up to and including $\tau-1$. By the equation of motion for the ARMA, 
conditional on this history, the residual variation will only be the unrealized $\varepsilon_\tau$ which is distributed normally. 

As previously argued, the history of $y$s along with a $y_0$ and $\varepsilon_0$
is sufficient to uniquely determine all of the $\varepsilon$s (up to $\tau-1$), so
we can include the recursively computed $\varepsilon_{\tau-1}$ in the probability density.

Since $y_{\tau-1}$ and $\varepsilon_{\tau-1}$ can be computed deterministically, and $\varepsilon_\tau$
is normally distributed, is a linear function of normal random variables, so each
term in the product will be a normal pdf with mean $\phi y_{\tau-1} + \gamma \varepsilon_{\tau-1}$ (from the realized portion) and variance $\sigma^2$ (from the unrealized portion).

The normal pdf of this form is given by:

$$
p(y_\tau|y_0,\varepsilon_0,y_{\tau-1},\ldots)=\frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(y_\tau-(\phi y_{\tau-1} + \gamma \varepsilon_{\tau-1}))^2}{2\sigma^2}\right)
$$
And then we can simplify our expression as:

$$
\begin{aligned}
p(\mathcal{I}_t|y_0,\varepsilon_0)&=\prod_{\tau=1}^tp(y_\tau|y_0,\varepsilon_0,y_{\tau-1},\ldots) \\
&=\prod_{\tau=1}^t\frac{1}{\sqrt{2\pi\sigma^2}}\exp\left(-\frac{(y_\tau-(\phi y_{\tau-1} + \gamma \varepsilon_{\tau-1}))^2}{2\sigma^2}\right) \\
&=\left(\frac{1}{\sqrt{2\pi\sigma^2}}\right)^t\prod_{\tau=1}^t\exp\left(-\frac{(y_\tau-(\phi y_{\tau-1} + \gamma \varepsilon_{\tau-1}))^2}{2\sigma^2}\right) \\
&=\left(\frac{1}{\sqrt{2\pi\sigma^2}}\right)^t\exp\left(\sum_{\tau=1}^t-\frac{(y_\tau-(\phi y_{\tau-1} + \gamma \varepsilon_{\tau-1}))^2}{2\sigma^2}\right) \\
&=\left(\frac{1}{\sqrt{2\pi\sigma^2}}\right)^t\exp\left(-\frac{1}{2\sigma^2}\sum_{\tau=1}^t(\underbrace{y_\tau-(\phi y_{\tau-1} + \gamma \varepsilon_{\tau-1})}_{\varepsilon_\tau})^2\right) \\
&=\left(\frac{1}{\sqrt{2\pi\sigma^2}}\right)^t\exp\left(-\frac{1}{2\sigma^2}\sum_{\tau=1}^t\varepsilon_\tau^2\right) \\
&=(2\pi\sigma^2)^{-t/2}\exp\left(-\frac{1}{2\sigma^2}\sum_{\tau=1}^t\varepsilon_\tau^2\right) \\
\end{aligned}
$$

Where those $\varepsilon_\tau$s can be compute deterministically from the data
according to the procedure in the previous question.

## 4. State Space Representation

A state space representation is a system of vector equations of the form:

$$
\begin{aligned}
y_t&=\mu+Az_t+H\eta_t\\
z_t&=\Phi z_{t-1}+R\xi_t
\end{aligned}
$$

Where $\eta_t\sim \mathcal{N}(0,I),\xi_t\sim\mathcal{N}(0,\Sigma)$. Letting $z_{t}=\begin{bmatrix} y_t & \varepsilon_t \end{bmatrix}'$ we can represent the observation equation as:

$$
y_t = \begin{bmatrix} 1 & 0 \end{bmatrix}\begin{bmatrix} y_t \\ \varepsilon_t\end{bmatrix}+0\eta_t
$$

The $\eta$ component will not be needed, so we can set $H=0$ . Here's a canonical
way to represent the state equation that ensures dimensional conformability at every step.

$$
\begin{bmatrix} y_t \\ \varepsilon_t \end{bmatrix} =
\begin{bmatrix} \phi&\gamma \\ 0&0
\end{bmatrix}
\begin{bmatrix} y_{t-1} \\ \varepsilon_{t-1} \\\end{bmatrix} +
\begin{bmatrix}1&0 \\ 1&0\end{bmatrix} \xi_t
$$

Where $\xi_t$ is a 2-long vector distributed $\xi_t\sim\mathcal{N}(0,\sigma^2 I)$.
You could also use a one dimensional $\xi_t$ and then just the first column of the
matrix before $\xi_t$.

## 5. Simulation

Yes, this process is covariance stationary because $\phi = 0.9 < 1$, which was sufficient to get the geometric series to converge from part 1 of the problem.

```{r}
# Parameters
phi <- 0.9
gamma1 <- 0.1
sig2 <- 1

theta <- c(phi, gamma1, sig2)
```

```{r}
simulateARMA <- function(t, theta, y0, eps0) {
  # ************************************************************************ #
  # Simulates ARMA(1,1) data of length t: 
  #   y_t = \phi y_{t-1} + \epsilon_t + \gamma_1 \epsilon_{t-1}, 
  # where \epsilon_t ~ N(0, \sigma^2) and \epsilon_0 = eps0, y_0 = y0
  #
  # Inputs:
  #   t: length of time series to simulate
  #   theta: model parameters phi, gamma1, sig2
  #   y0, eps0: initial value of y_0 and \epsilon_0
  #
  # Output: List
  #   y: vector of simulated data y = (y_1, ... y_t)
  #   eps: vector of shocks
  # ************************************************************************ #

  phi <- theta[1]
  gamma1 <- theta[2]
  sig2 <- theta[3]
  
  eps <- c(eps0, rnorm(t, mean = 0, sd = sqrt(sig2)))  # draw t shocks
  
  y <- numeric(t+1)  # init zero vector of length t
  y[1] <- y0         # set y_0 = y0
  for (i in 2:(t+1)) {
    y[i] <- phi * y[i - 1] + eps[i] + gamma1 * eps[i - 1]
  }
  return(list(y = y[-1], eps = eps[-1]))  # return y_1,...,y_t and eps_1,...,eps_t
}
```

```{r}
set.seed(seed_number)

# Simulate data with t=150, y0=0, eps0=0
t <- 150
sim_data <- simulateARMA(t, theta, y0 = 0, eps0 = 0)
y_sim <- sim_data$y
eps_sim <- sim_data$eps

# Plot the simulated data
plot(y_sim, type = "l", col = "blue", lwd = 2,
     main = "Simulated ARMA(1,1) Data", xlab = "Time", ylab = "y_t")
```

## 6. Implementing the Kalman Filter

Letting $v_i = y_i - A\hat{z}_{i|i-1}$ be the prediction error of the Kalman filter and $F_i := \mathbb{E}[v_i^2 \mid Y^{i-1}] = AP_{i|i-1}A'$ be the forecast variance, the Kalman filter likelihood is given by
$$
p_\text{KF}(\mathcal{I}^t) = \prod_{i=1}^t \frac{1}{\sqrt{2\pi F_i}} \exp\left( -\frac{v_i^2}{2F_i} \right)
$$
The analytical expression from part 3 is
$$
p_\text{A}(\mathcal{I}^t \mid y_0, \varepsilon_0) = \prod_{i=1}^t \frac{1}{\sqrt{2\pi \sigma^2}} \exp\left( -\frac{\varepsilon_i^2}{2\sigma^2} \right)
$$
Comparing the two, we see that $v_i$ and $F_i = \mathbb{E}[v_i^2 \mid Y^{i-1}]$ are the Kalman equivalent of $\varepsilon_i$ and $\sigma^2 = \mathbb{E}[\varepsilon_i^2]$, respectively. This makes sense because the prediction error $v_i$ can be thought of as the Kalman filter's estimate of the shock $\varepsilon_i$ at time $i$, and so the forecast variance $F_i$ can be thought of as the Kalman filter's estimate of the variance of that shock.

For our simulated data, the log-likelihood computed by the Kalman filter is very close to that computed by the analytical expression derived in part 3 (respectively, $-209.5759$ and $-209.3769$). Since the Kalman estimates of $y_t$ and $\varepsilon_t$ are essential exact matches to the simulated data (as discussed in part 7), it makes sense that the log-likelihoods are also very close.

```{r}
kalmanFilter <- function(y, theta) {
  # ************************************************************************ #
  # Implements Kalman filter for ARMA(1,1) model:
  #   y_t = \phi y_{t-1} + \epsilon_t + \gamma_1 \epsilon_{t-1},
  # where \epsilon_t ~ N(0, \sigma^2). 
  # We use state vector s_t = (y_t, eps_t).
  #
  # Inputs:
  #   y: vector of observed data (length t)
  #   theta: model parameters phi, gamma1, sig2
  # 
  # Output: List
  #   s: vector of state estimates s_t = E[s_t | y^t]
  #   P: vector of state covariance matrices P_t = Var[s_t | y^t]
  #   logLik: log likelihood of data: log L(y) = \sum_{t=1}^T log p(y_t | y^{t-1})
  # ************************************************************************ #
  
  phi <- theta[1]
  gamma1 <- theta[2]
  sig2 <- theta[3]
  
  t <- length(y)
  s <- array(0, dim = c(t, 2))     # init state estimates
  colnames(s) <- c("y", "eps")
  P <- array(0, dim = c(2, 2, t))  # init state covariance matrices
  
  # State space representation
  A   <- matrix(c(1,0), nrow = 1, ncol = 2)
  Phi <- matrix(c(phi, gamma1, 0, 0), nrow = 2, ncol = 2, byrow = TRUE)
  R   <- matrix(c(1, 1), nrow = 2, ncol = 1)

  # Initialization using unconditional moments (s_{1|0}, P_{1|0})
  s[1, ] <- c(0, 0)
  p11 <- (1 + (phi + gamma1)^2 / (1 - phi^2)) * sig2  # derived in part 1
  P[, , 1] <- matrix(c(p11, sig2, sig2, sig2), nrow = 2, ncol = 2)

  # Initialize log-likelihood
  logLik <- 0

  # Kalman filter recursion
  for (i in 1:t) {
    # Prediction
    v <- as.numeric(y[i] - A %*% s[i, ])        # y_i - As_{i|i-1}
    Evv <- as.numeric(A %*% P[, , i] %*% t(A))  # AP_{i|i-1}A'
    
    # Add log-likelihood
    logLik <- logLik - 0.5 * (log(2 * pi) + log(Evv) + (v^2) / Evv)
    
    # Update: s_{i|i} and P_{i|i}
    s[i, ] <- s[i, ] + P[, , i] %*% t(A) * v / Evv
    P[, , i] <- P[, , i] - (P[, , i] %*% t(A) %*% A %*% t(P[, , i])) / Evv
     
    # Predict: s_{i+1|i} and P_{i+1|i}
    if (i < t) {
      s[i+1, ] <- Phi %*% s[i, ]
      P[, , i+1] <- Phi %*% P[, , i] %*% t(Phi) + R %*% sig2 %*% t(R)
    }
  }
  
  return(list(s = s, P = P, logLik = logLik))
}
```

```{r}
logLik_analytical <- function(y, eps, theta, y0, eps0) {
  # ************************************************************************ #
  # Evaluates the log-likelihood of the data given the shocks, 
  # using the analytical expression derived in part 3.
  #
  # Inputs:
  #   y: vector of observed data (length t)
  #   eps: vector of shocks (length t)
  #   theta: model parameters phi, gamma1, sig2
  #   y0, eps0: initial values of y_0 and eps_0
  #
  # Output:
  #   log_lik: log-likelihood of data given eps
  # ************************************************************************ #
  
  t <- length(y)
  sig2 <- theta[3]
  log_lik <- -0.5 * t * log(2 * pi * sig2) - (1 / (2 * sig2)) * sum(eps^2)
  return(log_lik)
}
```


## 7. Applying the Kalman Filter

For $y_t$, the Kalman filter estimates exactly match the observed values. In our state space representation, the observation equation $y_t = Az_t$ has no noise term, so the data is perfectly informative about the latent state and the filter perfectly recovers the $y_t$ component of the state vector. 

For $\varepsilon_t$, the Kalman filter estimates also seem to perfectly match the observed shocks, except for the first few periods. This is because the initial state estimate for $\varepsilon_1$ is 0, which is not correct. However, as the filter processes more data, it quickly learns the true values of $\varepsilon_t$ and the estimates converge to the true shocks.

```{r}
# Run Kalman filter on simulated data
kf_result <- kalmanFilter(y_sim, theta)
s_est <- kf_result$s
P_est <- kf_result$P
logLik <- kf_result$logLik
logLik_analytical <- logLik_analytical(y_sim, eps_sim, theta, y0 = 0, eps0 = 0)
cat("Kalman Filter Log-Likelihood:", logLik, "\n",
    "Analytical Log-Likelihood:", logLik_analytical, "\n",
    sep = "")

# Plot state estimates alongside data
plot(y_sim, type = "l", col = "blue", lwd = 2,
     main = "Kalman Filter State Estimates: y_t", xlab = "Time", ylab = "y_t")
lines(s_est[, "y"], col = "red", lwd = 2)
legend("topright", legend = c("Observed", "Estimated"), 
       col = c("blue", "red"), lwd = 2)

plot(eps_sim, type = "l", col = "blue", lwd = 2,
     main = "Kalman Filter State Estimates: epsilon_t", xlab = "Time", ylab = "eps_t")
lines(s_est[, "eps"], col = "red", lwd = 2)
legend("topright", legend = c("Observed", "Estimated"), 
       col = c("blue", "red"), lwd = 2)
```

# Metropolis Hastings

## 1. Numerical Maximization

Because we are using uniform (flat) priors, the posterior is a constant times the likelihood. Therefore, maximizing the posterior is equivalent to maximizing the likelihood. 

```{r}
computeLogLik <- function(theta, y) {
  # ************************************************************************ #
  # Evaluates and maximizes the posterior
  #   log L(y^T | theta) + log(p(theta)) 
  # where p(theta) is the prior over theta = (phi, gamma1, sig2). 
  # Assuming flat priors, this reduces to maximizing log L(y^T | theta)
  #
  # Inputs:
  #   theta: model parameters phi, gamma1, sig2
  #   y: vector of observed data (length t)
  #
  # Output:
  #   log_lik: log-likelihood of data given theta
  # ************************************************************************ #
  
  # Handles constraints on theta
  phi <- theta[1]
  sig2 <- theta[3]
  if (abs(phi) >= 1 || sig2 <= 0) {
    return(-Inf)
  }
  
  # Evaluate log-likelihood using Kalman filter
  kf_out <- kalmanFilter(y, theta)
  log_lik <- kf_out$logLik
  
  return(log_lik)
}

max_likelihood <- function(y) {
  # ************************************************************************ #
  # Maximizes the log-likelihood. See computeLogLik().
  #
  # Inputs:
  #   y: vector of observed data (length t)
  #
  # Output: 
  #   theta_bar: parameter values that maximize the log-likelihood
  #   Sigma: inverse of negative Hessian of log-likelihood at theta_bar
  # ************************************************************************ #
  
  # Initial guess for theta = c(phi, gamma1, sig2)
  theta_init <- c(phi = 0, gamma1 = 0.1, sig2 = var(y))
  
  # Run optimization
  opt_result <- optim(
    par = theta_init,
    fn = computeLogLik,
    y = y,
    method = "L-BFGS-B",  # Broydon's method :)
    lower = c(-0.99, -Inf, 0.01),
    upper = c(0.99, Inf, Inf),
    control = list(fnscale = -1),  # minimize negative LL
    hessian = TRUE
  )
  
  theta_bar <- opt_result$par
  negH <- -opt_result$hessian
  Sigma <- solve(negH)
  
  return(list(theta_bar = theta_bar, Sigma = Sigma))
}
```

```{r}
# Estimate parameters using MLE
mle_result <- max_likelihood(y_sim)

theta_bar <- mle_result$theta_bar
Sigma <- mle_result$Sigma
```

## 2. Generating M-H Draws

```{r}
metropolisHastings <- function(y, theta_bar, Sigma, N = 10000, c = 1) {
  # ************************************************************************ #
  # Implements Metropolis-Hastings algorithm.
  #
  # Inputs:
  #   y: vector of observed data
  #   theta_bar: parameter values maximizing log-likelihood
  #   Sigma: inverse of negative Hessian of log-likelihood at theta_bar
  #   N: number of MCMC samples to draw
  #   c: scales variance in candidate density
  #
  # Output:
  #   theta_samples: matrix of sampled parameter values (N x length(theta))
  # ************************************************************************ #
  theta_samples <- matrix(NA, nrow = N, ncol = length(theta_bar))
  colnames(theta_samples) <- c("phi", "gamma1", "sig2")
  
  # Counter for acceptance rate
  accept_count <- 0
  
  # Draw initial sample
  theta_prev <- rmvnorm(n = 1, mean = theta_bar, sigma = c*Sigma)
  theta_samples[1,] <- theta_prev
  logLik_prev <- computeLogLik(theta_prev, y)
  
  # theta_prev stores theta_{i-1}
  # logLik_prev stores logL(y|theta_{i-1})
  
  for (t in 2:N) {
    theta_cand <- rmvnorm(n = 1, mean = theta_prev, sigma = c*Sigma)
    # logLik_curr stores logL(y|theta_cand)
    logLik_curr <- computeLogLik(theta_cand, y)
    alpha <- min(1, logLik_curr - logLik_prev)  # log transformed
    if (is.finite(logLik_curr) && (log(runif(1)) < alpha)) {
      # is.finite() rejects infeasible candidates (e.g., |phi|>1)
      theta_samples[t,] <- theta_cand
      theta_prev <- theta_cand
      logLik_prev <- logLik_curr
      accept_count <- accept_count + 1
    } else {
      theta_samples[t,] <- theta_prev
    }
    
    if (t %% (N / 10) == 0) {
      cat(sprintf("Iteration %d / %d | Current Acceptance Rate: %.2f%%\n", 
                  t, N, (accept_count / t) * 100))
    }
  }
  
  cat(sprintf("\nFinal Acceptance Rate: %.2f%%\n", (accept_count / N) * 100))
  return(theta_samples)
}
```

## 3. Lots of Simulations

```{r lots-of-simulation-be-patient}
set.seed(seed_number)
N <- 10000  # CHANGE TO 500000 BEFORE SUBMITTING
MHdraws <- metropolisHastings(y_sim, theta_bar, Sigma, N = N, c = 2)
```

```{r}
# Discard burn-in
burn_in <- 0.6 * N
MHdraws_conv <- MHdraws[(burn_in + 1):N, ]  # converged MH draws

post_stats <- apply(MHdraws_conv, 2, function(x) {
  density_est <- density(x)
  c(
    mean     = mean(x),
    median   = median(x),
    mode     = density_est$x[which.max(density_est$y)],
    ci_lower = quantile(x, 0.025),
    ci_upper = quantile(x, 0.975)
  )
}) %>% 
  t() %>%
  as.data.frame() %>% 
  mutate(parameter = rownames(.)) %>%
  relocate(parameter)
post_stats
```

Our marginal posterior densities are centered closely around the true parameter values. The posterior modes for $\phi$, $\gamma_1$, and $\sigma^2$ are respectively `r round(post_stats$mode[1], 3)`, `r round(post_stats$mode[2], 3)`, and `r round(post_stats$mode[3], 3)`, which are close to the true values $\phi = 0.9$, $\gamma_1 = 0.1$, and $\sigma^2 = 1$. The distributions are also relatively tight. The 95\% credible intervals are:

* $\phi$: [`r round(post_stats$ci_lower[1], 3)`, `r round(post_stats$ci_upper[1], 3)`]
* $\gamma_1$: [`r round(post_stats$ci_lower[2], 3)`, `r round(post_stats$ci_upper[2], 3)`]
* $\sigma^2$: [`r round(post_stats$ci_lower[3], 3)`, `r round(post_stats$ci_upper[3], 3)`]

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)

# Plot marginal density of params
true_vals <- data.frame(
  parameter = factor(c("phi", "gamma[1]", "sigma^2"), 
                     levels = c("phi", "gamma[1]", "sigma^2")),
  true_value = c(0.9, 0.1, 1)
)
draws_df <- as.data.frame(MHdraws_conv) %>%
  pivot_longer(cols = c("phi", "gamma1", "sig2"), 
               names_to = "parameter", 
               values_to = "value") %>%
  mutate(parameter = factor(parameter, 
                            levels = c("phi", "gamma1", "sig2"),
                            labels = c("phi", "gamma[1]", "sigma^2")))

ggplot(draws_df, aes(x = value)) +
  geom_density(fill = "steelblue", alpha = 0.4, color = "navy", linewidth = 0.8) +
  geom_vline(data = true_vals, aes(xintercept = true_value), 
             color = "firebrick", linetype = "dashed", linewidth = 1) +
  facet_wrap(~ parameter, scales = "free", ncol = 1, labeller = label_parsed) +
  labs(title = "Marginal Posterior Densities",
       x = "Parameter Value",
       y = "Density") +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    strip.background = element_rect(fill = "grey95", color = "grey50"),
    strip.text = element_text(size = 14),
    panel.grid.minor = element_blank()
  )
```
